\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{arev}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{braket}
\usepackage{bytefield}
\usepackage{qcircuit}

\usepackage{minted}
\usepackage{mdframed}
\usepackage{mathtools}
\usepackage[colorlinks=true,allcolors=black]{hyperref}
%\usepackage{authblk}

\surroundwithmdframed{minted}

\newminted[quil]{text}{fontsize=\footnotesize}
\newminted[python]{python}{fontsize=\footnotesize}
\newminted[bash]{bash}{fontsize=\footnotesize}


\newcommand{\RR}{\ensuremath{\mathbb{R}}}
\newcommand{\CC}{\ensuremath{\mathbb{C}}}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\re}{Re}
\DeclareMathOperator{\im}{Im}
\DeclareMathOperator{\image}{image}
\DeclareMathOperator{\Run}{\mathcal{R}}

\newcommand{\Zero}{\ensuremath{\ket{0}}}
\newcommand{\One}{\ensuremath{\ket{1}}}

\newcommand{\RX}{\ensuremath{\mathtt{RX}}}
\newcommand{\RY}{\ensuremath{\mathtt{RY}}}
\newcommand{\RZ}{\ensuremath{\mathtt{RZ}}}
\newcommand{\CZ}{\ensuremath{\mathtt{CZ}}}
\newcommand{\TGate}{\ensuremath{\RZ(\tfrac{\pi}{4})}}
\newcommand{\TdGate}{\ensuremath{\RZ(-\tfrac{\pi}{4})}}
\newcommand{\CNOT}{\ensuremath{\mathtt{CNOT}}}
\newcommand{\SWAP}{\ensuremath{\mathtt{SWAP}}}
\newcommand{\Hadamard}{\ensuremath{\mathtt{H}}}
\newcommand{\PauliX}{\ensuremath{\mathtt{X}}}
\newcommand{\PauliY}{\ensuremath{\mathtt{Y}}}
\newcommand{\PauliZ}{\ensuremath{\mathtt{Z}}}
\newcommand{\Halt}{\ensuremath{\mathtt{HALT}}}

\title{CS239 Lecture Notes (Week 6)}
\author{%
Robert S.\ Smith\footnote{email: \href{mailto:robert@rigetti.com}{\texttt{robert@rigetti.com}} (corresponding author)} \and
Mark G.\ Skilbeck\footnote{email: \href{mailto:mark.skilbeck@rigetti.com}{\texttt{mark.skilbeck@rigetti.com}}}%
}
\date{}

\begin{document}

\maketitle

\begin{center}
\textit{These are notes for the 6 May 2019 lecture for UCLA's CS239 course. Questions, corrections, and comments are welcome!}
\end{center}

\tableofcontents

% Robert's section
\input{part1.tex}

\newpage

% Mark's section
\input{part2.tex}

\addcontentsline{toc}{section}{References}
\begin{thebibliography}{1}
  \bibitem{quilpaper} Robert S. Smith, Michael J. Curtis, William J. Zeng. {\em A Practical Quantum Instruction Set Architecture}. arXiv.org preprint. arXiv:1608.03355. 2016.
  \bibitem{noisycomp} Eric C. Peterson, Gavin E. Crooks, Robert S. Smith. {\em Fixed-Depth Two-Qubit Circuits and the Monodromy Polytope} arXiv.org preprint. arXiv:1904.10541. 2019.
  \bibitem{shouts} Robert S. Smith. {\em Someone Shouts $\ket{01000}$! Who's Excited?} arXiv.org preprint. arXiv:1711.02086. 2017.
  \bibitem{skalg} Christopher M. Dawson, Michael A. Nielson. {\em The Solovay--Kitaev Algorithm}. Quantum Information and Computation. 2005. 
  \bibitem{molmer} Klaus M{\o}lmer, Anders S{\o}rensen. {\em Multi-particle entanglement of hot trapped ions}. Physical Review Letters 82. 1999.
  \bibitem{shende} Vivek V. Shende, Igor L. Markov. {\em On the CNOT-cost of TOFFOLI gates}. Quantum Information and Computation. 2009.
  \bibitem{qft} Craig Gidney. {\it Building your own Quantum Fourier Transform}. Blog post: \url{https://algassert.com/quantum/2014/03/07/Building-your-own-Quantum-Fourier-Transform.html}. 2014. (Accessed 6 May 2019)
\end{thebibliography}

\end{document}
